---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ppa

<!-- badges: start -->
<!-- badges: end -->

The goal of ppa is to provide an useable example implementation of plot panel analysis.
Plot panel analysis is a visualization technique that can be used to rapidly explore and analyze a
dataset when the observations of interest are complex objects like time series or scatter plots,
rather than the individual data points that comprise those objects.  Plot panel analysis is especially important
when no additional variables exist that can help you organize and make sense of the variation across the complex
objects.

## Installation

You can install the development version of ppa from [GitHub](https://github.com/) with:


``` r
# install.packages("devtools")
devtools::install_github("daattali/shinyalert") #requires dev version of shinyalert
devtools::install_github("jrodu/ppa")
```

Please note that at present installing "shinyalert" from Github is required, as there is a bug fix in that package that has not yet been pushed to CRAN.

## A simple motivational example

Consider the following set of time series

```{r oracle, echo=TRUE, message=FALSE, out.width='30%'}
library(tidyverse)
library(ppa)

set.seed(140)

#generate 4 different signals
tsa <- 1:50
tsb <- c(rep(1, 25), rep(50, 25))
tsc <- c(seq(50 ,41, length.out=20), seq(40, 11, length.out=10), seq(10,1, length.out=20))
tsd <- rep(25, 50)

data <- tibble(seconds = rep(1:50, times=100),
               haty=c(rep(tsa, times=25),
                      rep(tsb, times=25),
                      rep(tsc, times=25), rnorm(1250, 25, 15)),
               series=c(rep(sample(1:100), each=50)),
               identity=c(rep(c('a', 'b', 'c', 'd'), each=1250)),
               noise=c(rnorm(5000, sd=15)),
               y=haty+noise) %>% arrange(series)

corrupt_one_series <- sample(1:100, 1)
data <- data %>% mutate(y=ifelse(series==corrupt_one_series, haty, y))

data %>% ggplot(aes(seconds, y))+geom_line(aes(group=series)) + facet_wrap(~identity) +
     xlab(NULL) +
     ylab(NULL) +
     theme(axis.text.x = element_blank(), axis.title.y = element_blank(),
           axis.text.y = element_blank(),
           axis.title.x = element_blank(),
           panel.border = element_blank(), panel.background = element_rect(fill="white"),
           panel.grid = element_line(colour = "grey70"))


```

In the previous figure, the signal is relatively clear.  However, what if we did not have access to the grouping labels?

```{r noisy, echo=FALSE, message=FALSE,fig.show="hold", out.width='30%'}
data %>% ggplot(aes(seconds, y))+geom_line(aes(group=series)) + 
xlab(NULL) +
  ylab(NULL) +
  theme(axis.text.x = element_blank(), axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        panel.border = element_blank(), panel.background = element_rect(fill="white"),
        panel.grid = element_line(colour = "grey70"), plot.margin=margin(0,2,0,0, "cm"))


```

Analysis would be difficult.  No additional variables are available
that could help the analyst easily pull apart the time series in a useful way.
Plot panel analysis can help with this!

**But before we delve in to how plot panel analysis can help identify signal across time series, let's
look at a much simpler problem that can be solved incredibly fast with plot panel analysis...**
Embedded in each of these figures (including the first figure) is one "degenerate"
time series that has no added noise.  Can you identify it?  Would you have known to look for it had you not seen the data generating code?  If you could identify them visually, could you
quickly locate them in the dataset and remove them from analysis?

This is easy to do with plot panel analysis (and using the ppa app).  Here are the steps.

1. Copy the code that generates the data above and paste and run it in your console.  You should now have a variable called "data" in your environment.  Run following code in your console
```{r ppaapp, eval=FALSE}
data %>% ppa(seconds, y, series, 'line')
```
2. locate the degenerate time series and click on it.  It should turn grey.
3. Click on the "Exclude/Unexclude Panels" checkbox.
4. Click on the "exclude" button.  The degenerate time series will disappear.
5. Exit the app by closing the window.
6. A vector called "excluded_panels" will show up in your environment which contains the 
unique panel id of the degenerate time series (this is the `series` variable in the original dataset)

**Simple!**

## Some features of the PPA app that are useful for plot panel analysis

Let's take a look at some of the features of the PPA app.  It's easiest to read this list while playing around with the app, so here is an example you can run.

```{r ppaappOtherExamples, eval=FALSE}
#example from above
data %>% ppa(seconds, y, series, 'line')
```

1.  Click on the "panel plot properties" button to
    + change the number of columns
    + change the height of the panels (expressed as a multiplier of the default size).
    The multiplier need not be an integer.  The size of the panels is relative
    to the window size.
2. The "Individual Panel Settings" checkbox
    + shows the x variable, the y variable, and the additional variables available in the dataset
    + allows you to toggle between plot types (currently "line" or "scatter")
3. The "Panel Shufflers" checkbox allows you to shuffle the panels either by 
shuffling the rows, shuffling the columns, shuffling the rows and columns (panels in the same row will stay
in the same row, panels in the same column will stay in the same column, but the location of the rows/columns will change), or shuffling by panel without respect to row or column.
4. The "Panel Selection Actions" checkbox allows you to
    + unselect currently selected panels (Unselect All)
    + zoom in on currently selected panels (Examine Selected)
    + undo the "zoom in" (Renturn to all panels)
5. The "Exclude/Unexclude Panels" checkbox allows you to exclude selected panels (removes them from the plot) and
unexclude all previously excluded panels.
6. The "Label Panels and Select by Label" checkbox allows you to label panels (you first must "add" a label (or select a previously added label) and then apply it to the selected panels) and select panels (that have been previously labeled) by choosing a label or set of labels.
7. The "Select Panels by Score Function" checkbox allows you to
    + create a feature function to "measure" each time series based on your visual observations
    + edit a feature function you already created (this is useful when you realize that you made a typo or that your feature function is not capturing what you thought it would).
    + score each time series based on your function
    + select panels by moving the threshold on the resulting ECDF plot
    + invert the selected panels (i.e. selecting panels that lie "above" the threshold rather than "below")
    + investigate the score of individual time series by hovering over them and exposing a tooltip.
    + save the scores to your R environment (the scores are saved in a data set whose columns are named based on the name of the function that produced the scores)
    + the functions you created will be saved to your environment as immediately usable functions, along with a list that contains the name of the functions created during that session.
    
Here is a feature function you can try (note- don't forget to name the function so that you can access it later!):
```{r ppaapp_score, eval=FALSE}
panel_data %>% summarize(max=max(y)) %>% pull(max)

#or (same thing, different syntax)

max(panel_data$y)

#you can copy and paste either of these code snippets into the popup after you click on create new function
```

8. The "Select Panels by Comparison" checkbox allows you to do the same as selecting by score function, but with a user-suppled function that compares each panel to a selected panel.
    + When scores are saved using this functionality, since the score depends on the "base comparison panel," the column name for the scores is in the format "name-of-function_base-comparison-panel-id"
9. The "Arrange Panels in Scatter Plot" checkbox allows you to observe the relationship between two
different measurement functions from the "Select by Score Function" module.
10. The "Transform Panels" checkbox allows you to rapidly investigate a data transformation or series of transformations.
    + When the first transformation is applied, a plot will appear in the right-hand column that is a tree plot that records the transformations attempted.  Multiple transformations in succession will result in a series of nodes corresponding to a "branch" of the tree.  
    + To revert to a previous transformation (either to continue with analysis or create a new branch of successive transformations) simply click on the node on the tree corresponding to your desired transformation "state."  The plot panels will revert to that state, and you can apply the desired transformation (or continue your investigation) from there.
    
    
Here is a transformation you can try (note- don't forget to name the function so that you can access it later!):
```{r ppaapp_transformation, eval=FALSE}
panel_data %>% mutate(y=y^2)
```
Go ahead, try a few transformations!  Try to chain them together in different ways and navigate through the various transformations to get a feel for using the navigation tree.

## Tip

Have a set of score/compare/transform functions that you know a prior you would like to use but don't want to re-type them into the function input boxes? No problem! Just make sure they are in your environment when you launch the app, and write a wrapper function for them in the input boxes in the app!
